Inferring the effective start dates of non-pharmaceutical interventions during \covid outbreaks

### 2020-05-24 ms/ms.tex
for at, revise, describe WAIC for uniform prior

We compared the posterior predictive plots of a model with a free $\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e. transmission and reporting rates are constant). The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}). 
When we compared the models using WAIC~(\autoref{table:WAIC}), the model with a free~$\tau$ was preferred in 8 out of 12 of the regions (although only narrowly for 5 of the 8).
The exceptions are Austria, Belgium, Norway, and United Kingdom.

We compared the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and found that in most regions the effective start of NPIs  significantly differs from the official date (\autoref{fig:tau-summary}), that is, the credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:tau-summary}). The exceptions are, as with the comparison to the simpler models, Austria, Belgium, and United Kingdom, as well as Switzerland (see below). Norway also has a relatively wide credible interval, maybe because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
In the following, we describe our findings in more detail.

%%% Late
\paragraph*{Late effective start of NPIs.}
In both Wuhan, China, and in Italy we find that our estimated effective start of NPI $\hat{\tau}$ is significantly later than the official date $\tau^*$.

@@

%%% Fig - Late %%%
\begin{figure}[h]
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/Italy_τ_posterior.pdf}
    \end{subfigure}
  	~
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/Wuhan_τ_posterior.pdf}
    \end{subfigure}
    \caption{
	\textbf{Late effect of non-pharmaceutical interventions in Italy and in Wuhan, China.}
    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimate $\hat{\tau}$. Shaded area show a 95\% credible interval (interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
    }
    \label{fig:late}
\end{figure}

@@

%%% Fig - Prediction %%%
\begin{figure}[h]
    \centering
	\includegraphics[width=0.5\textwidth]{../figures/Fig-prediction.pdf}
    \caption{
    \textbf{Late effective start of NPIs leads to underestimation of daily confirmed cases.}
    Real number of daily cases in Italy in black (markers: data, line: time moving average). 
    Model posterior predictions are shown as coloured lines (1,000 draws from the posterior distribution). 
    Shaded box illustrates a serial interval of seven days.
    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model underestimates the number of cases seven days after the start of the NPI.
    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
    Here, model parameters are best estimates for Italy (\autoref{table:estimated-params}).
    } 
    \label{fig:prediction}
\end{figure}

@@

For example, in Italy and Wuhan, China, the effective start of the lockdowns seems to have occurred five days or more after the official date (\autoref{fig:late}).
This difference might be explained by low compliance: In Italy, for example, the government intention to lockdown Northern provinces leaked to the public, resulting in people leaving those provinces~\citep{Gatto2020}.
Late effect of NPIs may also be due to the time required by both the government and the citizens to organise for a lockdown, and for the new guidelines to diffuse in the population.

@@

We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in 9 out of 12 regions (i.e. free or fixed model in \autoref{table:WAIC}).
It could be interesting to check if the evidence favours a model with \emph{two} change-points, rather than one. 
Two such change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
However, interpretation of such models will be harder, as two change-points can also reflect a mix of NPIs and other events, such as changing weather, news of new treatments, and international outbreaks.

@@

\paragraph*{Conclusions.}
We have inferred the effective start date of NPIs and found that they often differ from the official dates.
Our results highlight the complex interaction between personal, regional, and global determinants of behavioral response to infectious disease.
Therefore, we emphasize the need to further study variability in compliance and behavior over both time and space. This can be accomplished both by surveying differences in compliance within and between populations~\citep{Atchison2020}, and by incorporating specific behavioral models into epidemiological models~\citep{Arthur2020,Fenichela2011,Walters2013}.

@@

We defined the following prior distributions on the model parameters $P(\theta)$: 
\begin{equation} \label{eq:priors}
\begin{aligned} 
Z & \sim \mathit{Uniform}(2, 5) \\
D & \sim \mathit{Uniform}(2, 5) \\
\mu & \sim \mathit{Uniform}(0.2, 1) \\
\beta & \sim \mathit{Uniform}(0.8, 1.5) \\
\lambda & \sim \mathit{Uniform}(0, 1) \\
\alpha_1, \alpha_2 & \sim \mathit{Uniform}(0.02, 1)\\
E(0) & \sim \mathit{Uniform}(0, 3000) \\
I_u(0) & \sim \mathit{Uniform}(0, 3000) \\
\Delta t & \sim \mathit{Uniform}(1, 5) \\
\tau &\sim \mathit{TruncatedNormal}\Big(\frac{\tau^*+\tau^0}{2}, \frac{\tau^*-\tau^0}{2}, 1, T-2\Big),
\end{aligned}
\end{equation}
where the prior for $\tau$ is a truncated normal distribution shaped so that the date of the first and last NPI, $\tau^0$ and $\tau^*$ (\autoref{table:NPI_dates}), are at minus and plus one standard deviation, and taking values only between 1 and  $T-2$, where $T$ is the number of days in the data $\vec{X}$.
We have also tested an uninformative uniform prior $\mathit{Uniform}(1,T-2)$. The uninformative prior could result in non-negligible posterior probability for unreasonable $\tau$ values, such as Mar~1 in the United~Kingdom. This was probably due to MCMC chains being stuck in low posterior regions of the parameter space. We therefore decided to use the more informative truncated normal prior.
Other priors follow \citet{Li2020}, with the following exceptions.
$\lambda$ is used to ensure transmission rates are lower after the start of the NPIs ($\lambda < 1$).
We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood and unreasonable parameter estimates, and therefore chose $\mathit{Uniform}(1,5)$ as the prior.

>>>>

We compare the posterior predictive plots of a model with a free $\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e. transmission and reporting rates are constant). The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}). 
When we compare the models using WAIC~(\autoref{eq:WAIC}, \autoref{table:WAIC}), the model with a free~$\tau$ is preferred in 8 out of 12 of the regions (although only narrowly for 5 of the 8).
The exceptions are Austria, Belgium, Norway, and United Kingdom.

We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPIs  significantly differs from the official date (\autoref{fig:tau-summary}), that is, the credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:tau-summary}).
The exceptions are, as with the comparison to the simpler models, Austria, Belgium, and United Kingdom, as well as Switzerland (see below). Norway also has a relatively wide credible interval, maybe because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
In the following, we describe our findings in more detail.

%%% Late
\paragraph*{Late effective start of NPIs.}
In both Wuhan, China, and in Italy we estimate that the effective start of NPIs $\hat{\tau}$ is significantly later than the official date $\tau^*$.

@@

%%% Fig - Late %%%
\begin{figure}[h]
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/Italy_τ_posterior.pdf}
    \end{subfigure}
  	~
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/Wuhan_τ_posterior.pdf}
    \end{subfigure}
    \caption{
	\textbf{Late effect of non-pharmaceutical interventions in Italy and in Wuhan, China.}
    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimate $\hat{\tau}$. Shaded area shows a 95\% credible interval (interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
    }
    \label{fig:late}
\end{figure}

@@

%%% Fig - Prediction %%%
\begin{figure}[h]
    \centering
	\includegraphics[width=0.4\textwidth]{../figures/Fig-prediction.pdf}
    \caption{
    \textbf{Late effective start of NPIs leads to underestimation of daily confirmed cases.}
    Real number of daily cases in Italy in black (markers: data, line: time moving average). 
    Model posterior predictions are shown as coloured lines (1,000 draws from the posterior distribution). 
    Shaded box illustrates a serial interval of seven days.
    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model underestimates the number of cases seven days after the start of the NPI.
    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
    Here, model parameters are best estimates for Italy (\autoref{table:estimated-params}).
    } 
    \label{fig:prediction}
\end{figure}

@@

For example, in Italy and in Wuhan, China, the effective start of the lockdowns seems to have occurred five days or more after the official date (\autoref{fig:late}).
This difference might be explained by low compliance: In Italy, for example, the government intention to lockdown Northern provinces leaked to the public, resulting in people leaving those provinces~\citep{Gatto2020}.
Late effect of NPIs may also be due to the time required by both the government and the citizens to organise for a lockdown, and for the new guidelines to be adopted by the population.

@@

We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in 9 out of 12 regions (i.e. free or fixed model in \autoref{table:WAIC}).
It could be interesting to check if the evidence favors a model with \emph{two} change-points, rather than one. 
Two such change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
However, interpretation of such models will be harder, as two change-points can also reflect a mix of NPIs and other events, such as changing weather, news of new treatments, and international outbreaks.

@@

\paragraph*{Conclusions.}
We have inferred the effective start date of NPIs and found that they often differ from the official dates.
Our results highlight the complex interaction between personal, regional, and global determinants of behavioral response to an epidemic.
Therefore, we emphasize the need to further study variability in compliance and behavior over both time and space. This can be accomplished both by surveying differences in compliance within and between populations~\citep{Atchison2020}, and by incorporating specific behavioral models into epidemiological models~\citep{Arthur2020,Fenichela2011,Walters2013}.

@@

We defined the following prior distributions on the model parameters $P(\theta)$: 
\begin{equation} \label{eq:priors}
\begin{aligned} 
Z & \sim \mathit{Uniform}(2, 5) \\
D & \sim \mathit{Uniform}(2, 5) \\
\mu & \sim \mathit{Uniform}(0.2, 1) \\
\beta & \sim \mathit{Uniform}(0.8, 1.5) \\
\lambda & \sim \mathit{Uniform}(0, 1) \\
\alpha_1, \alpha_2 & \sim \mathit{Uniform}(0.02, 1)\\
E(0) & \sim \mathit{Uniform}(0, 3000) \\
I_u(0) & \sim \mathit{Uniform}(0, 3000) \\
\Delta t & \sim \mathit{Uniform}(1, 5) \\
\tau &\sim \mathit{TruncatedNormal}\Big(\frac{\tau^*+\tau^0}{2}, \frac{\tau^*-\tau^0}{2}, 1, T-2\Big),
\end{aligned}
\end{equation}
where the prior for $\tau$ is a truncated normal distribution shaped so that the date of the first and last NPI, $\tau^0$ and $\tau^*$ (\autoref{table:NPI_dates}), are at minus and plus one standard deviation, and taking values only between 1 and  $T-2$, where $T$ is the number of days in the data $\vec{X}$.
We also tested an uninformative uniform prior, $\mathit{Uniform}(1,T-2)$.
WAIC (\autoref{eq:WAIC}) of a model with this uniform prior was either higher, or lower by less than 2, compared to WAIC of a model with the truncated normal prior.
The uninformative prior resulted in non-negligible posterior probability for unreasonable $\tau$ values, such as Mar~1 in the United~Kingdom. This was probably due to MCMC chains being stuck in low posterior regions of the parameter space.
We therefore decided to use the more informative truncated normal prior for $\tau$.
Other priors follow \citet{Li2020}, with the following exceptions.
$\lambda$ is used to ensure transmission rates are lower after the start of the NPIs ($\lambda < 1$).
We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood and unreasonable parameter estimates, and therefore chose $\mathit{Uniform}(1,5)$ as the prior for $\Delta t$.
